\LARGE \bf Expectation-Maximization based defense mechanism for distributed Model Predictive Control

### 2022-02-07 article/article.tex
correction Simon

The main objective of the \EM{} algorithm is to find maximum likelihood estimators for models with latent variables.

@@

For an observation $o$, we have the input and response variables, identified as  ${\random{\vec{\theta}}_{o}}$ and ${\random{\vec{\lambda}}_{o}}$.
As~\eqref{eq:linear_cheating} gives us a multidimensional \pwa{} function, we propose to use a expansion of the model referred as \emph{mixture of switching regressions} in~\cite{QuandtRamsey1978} and \emph{mixture of linear regressions} in~\cite{FariaSoromenho2010}, which we will call \emph{mixture of switching affine regressions}, since our regressors have a linear term (matrices $\tilde{P}^{z}$) and a constant term (vectors $\tilde{\vec{s}}^{z}$):
% TODO(accacio): change subscript n
  \begin{equation}
    \label{eq:linear_cheating_random}
    \random{\lambda}_{o}=
        \begin{cases}
      -\tilde{P}^{1}\randomvec{\theta}_{o}-\tilde{\vec{s}}^{1}&\text{with probability}\ \pi_{1} \\
      \qquad\quad \vdots&\qquad\quad \vdots\\
      -\tilde{P}^{Z}\randomvec{\theta}_{o}-\tilde{\vec{s}}^{Z}&\text{with probability}\ \pi_{Z} \\
    \end{cases}
  \end{equation}

As we can see, this observation does not gives us the complete data, since we don't know which zone generated the response, but only the probability of the input being in that zone.
We consider the complete data consisting of the input, the response, and a non-observable variable. Therefore, for each observation $o$ there exists a latent random variable ${\random{z}_{o}\in\set{Z}}$, with associated probabilities ${\Pi=\{\pi_{1},\dots,\pi_{Z}\}}$ such prior probability
   \begin{equation}
     \label{eq:prob_zo_equal_z}
\probability{\random{z}_{o}=z}=\pi_{z}.
\end{equation}
Since $\vec{\theta}$ is our input, we know \emph{almost surely} their values giving
\begin{equation}
  \label{eq:theta_almost_surely}
  \probability{\randomvec{\theta}_{o}}=1.
\end{equation}
With this information we can use the graph in Fig.~\ref{fig:model} to represent the model.
\begin{figure}[b]
  \centering
  \begin{tikzpicture}
    \draw[thick,blue,rounded corners=10pt] (-1.7,1.2) rectangle (1.7,-1.7);
    \node at (1.2,-1.3) {O};
    \node (pi) at (-2.5,.5) {$\Pi$};
    \node (phi) at (2.5,-1.) {$\set{P}$};

    \graph [edge quotes={fill=white,inner sep=1pt},
    clockwise=3,nodes={circle,draw,rotate=-60,minimum width=1cm}] {
      a/"${\randomvec{\theta}_{o}}$"[rotate=60],b/"${\random{\vec{\lambda}}_{o}}$"[rotate=60],c/"${\random{z}}_{o}$"[rotate=60];
      {a,c} ->[-latex] b;
      (pi) ->[thick,{Circle[length=3.pt]}-latex] c;
      (phi) ->[thick,{Circle[length=3.pt]}-latex] b;
      };
  \end{tikzpicture}
  \caption{Graph representation of model proposed.}\label{fig:model}
\end{figure}

As said, the objective of the \EM{} algorithm is to maximize the likelihood, more precisely the complete data log-likelihood $\ln\probability{\random{\Theta},\random{\Lambda},\random{Z};\set{P}}$.
Using the dependency graph in Fig.~\ref{fig:model} yields
    \begin{equation}\label{eq:completedataLogLikelihood}
      \ln\probability{\random{\Theta},\random{\Lambda},\random{Z};\set{P}}=\ln \prod_{o=1}^{O}\probability{\randomvec{\theta}_{o},\randomvec{\lambda}_{o},\random{z}_{o};\set{P}} .
    \end{equation}
where the probability $\probability{\randomvec{\theta}_{o},\randomvec{\lambda}_{o},\random{z}_{o};\set{P}}$ is
\begin{equation}\label{eq:condional}
 \prod_{z=1}^{Z}{[\probability{\randomvec{\lambda}_{o}|\randomvec{\theta}_{o},\random{z}_{o}=z;\set{P}^{z}}\probability{\random{z}_{o}=z}\probability{\randomvec{\theta}_{o}}]}^{\indicator{\random{z}_{o}=z}},
\end{equation}
and $\set{P}^{z}$ is the $z$-th tuple in $\set{P}$.

For the likelihood $\probability{\randomvec{\lambda}_{o}|\randomvec{\theta}_{o},\random{z}_{o}=z,\set{P}^{z}}$, we use multivariate normal probability density functions
\begin{equation}
  \label{eq:multivariate_gaussian}
\mathcal{N}(\randomvec{\lambda}_{o};f(\randomvec{\theta}_{o};\set{P}^{z}),{\Sigma^{z}}),
\end{equation}
where the mean is
\[f(\randomvec{\theta}_{o};(P,\vec{s}))=-{P}\randomvec{\theta}_{o}-{\vec{s}}\]
and the covariance is ${\Sigma^{z}}$, which tends to $0$.

Using~\eqref{eq:prob_zo_equal_z},~\eqref{eq:theta_almost_surely}, and~\eqref{eq:multivariate_gaussian} in~\eqref{eq:completedataLogLikelihood} yields
\begin{equation}\label{eq:completedataLogLikelihood_complete}
\ln\probability{\random{\Theta},\random{\Lambda},\random{Z};\set{P}}=  \sum_{o=1}^{O}\sum_{z=1}^{Z}{\indicator{\random{z}_{o}=z}}
  \alpha_{zo},
\end{equation}
where ${\alpha_{zo}=\ln{\pi_{z}}+\ln{\mathcal{N}(\randomvec{\lambda}_{o};f(\randomvec{\theta}_{o};\set{P}^{z}),{\Sigma^{z}})}}$.

@@

  Given a set of observed data $(\random{\Theta},\random{\Lambda})$, estimate the unknown latent variables $\random{Z}$ and parameter set $\set{P}$ which maximize the complete-data log likelihood in~\eqref{eq:completedataLogLikelihood_complete}.
\end{problem}

But as mentioned, we do not observe the complete data $(\random{\Theta},\random{\Lambda},\random{Z})$, instead we observe only $(\random{\Theta},\random{\Lambda})$ and all information we can get from $\random{Z}$ is the posterior probability  ${\zeta(z_{zo};\set{P})=\probability{\random{z}_{o}=z|\randomvec{\lambda}_{o},\randomvec{\theta}_{o};\set{P}}}$, also called \emph{responsibility}, which can be calculated as
\begin{align}
  \label{eq:responsibilities}
\zeta(z_{zo};\set{P})&=\frac{\probability{\random{z}_{o}=z}\probability{\randomvec{\lambda}_{o}|\randomvec{\theta}_{o};\set{P}^{z}}}{\sum\limits_{j=1}^{Z}\probability{\random{z}_{o}=j}\probability{\randomvec{\lambda}_{o}|\randomvec{\theta}_{o};\set{P}^{j}}}\nonumber\\
  &=\frac{\pi_{z}{\mathcal{N}(\randomvec{\lambda}_{o};f(\randomvec{\theta}_{o};\set{P}^{z}),{\Sigma^{z}})}}{\sum\limits_{j=1}^{Z}\pi_{j}\mathcal{N}(\randomvec{\lambda}_{o};f(\randomvec{\theta}_{o};\set{P}^{j}),{\Sigma^{j}})}.
\end{align}

Observe that by taking
\begin{equation}\label{eq:argmaxz}
\underset{z_{o}}{\argmax}\ {\zeta(z_{zo};\set{P})},
\end{equation}
we can get the most probable $z$-zone which generated the observation $\randomvec{\lambda}_{o}$.

So, rather than using the complete data log-likelihood~\eqref{eq:completedataLogLikelihood_complete}, which we do not have, we use its expectation with respect to the prior calculated using a given set of parameter estimates $\set{P}_{\mathrm{cur}}$,  ${Q\left(\set{P},\set{P}_{\mathrm{cur}}\right)=\expectation[{\zeta(z_{zo};\set{P}_{\mathrm{cur}}}]{\ln\probability{\random{\Theta},\random{\Lambda},\random{Z};\set{P}}}}$:
\begin{equation}
  \label{eq:completedataLogLikelihood_expectation}
 Q\left(\set{P},\set{P}_{\mathrm{cur}}\right) =
  \sum_{o=1}^{O}\sum_{z=1}^{Z}  \zeta(z_{zo};\set{P}_{\mathrm{cur}})\alpha_{zo}
      ,
\end{equation}

Then we can find a new estimate of $\set{P}$ that maximizes $Q(\set{P},\set{P}_{\mathrm{cur}})$:
\begin{equation} \label{eq:Mstep}
  \set{P}_{\mathrm{new}}=\underset{\set{P}}{\argmax}\ Q(\set{P},\set{P}^{\mathrm{cur}}).
\end{equation}

With this information we can describe the \EM{} algorithm in two parts, first calculate the responsibilities $\zeta(z_{zo};\set{P}_{\mathrm{cur}})$, and then update the parameters.
Algorithm~\ref{alg:em} adpated from~\mbox{\cite[Chapter 9]{Bishop2006}} describes the steps.

@@

We can see some similarities to the K-planes algorithm (see~\cite{BradleyMangasarian2000}), but \EM{} is more compromising.
Instead of affecting the observed data to a cluster with 100\% of certainty (\emph{hard assignment}), \EM{} uses the responsibilities (\emph{soft assignment}) for each cluster.
Then, when we solve the \textbf{M Step}, rather than using only the data assigned to update the parameters of a certain cluster, we take into account all data, whose contribution to the update is weighted by their associated responsibilities.

Here we introduce a variable ${\phi^{z}={[\vectorize{\tilde{P}^{z}}^{T}\ {(\tilde{\vec{s}}^{z})}^{T} ]}^{T}}$ so we can calculate the new estimates of $\set{P}$ from~\eqref{eq:Mstep}. Using the KKT conditions~\cite{BoydVandenberghe2004}, we can find an optimal solution for the problem in~\eqref{eq:Mstep}, by
taking the gradients of~\eqref{eq:completedataLogLikelihood_expectation} with respect to vectors $\vec{\phi}^{z}$ and making them vanish.
Because of the multidimensional nature of the problem, some matrix operations are needed to synthesize the results.
After those operations, we have a matricial solution that yields the optimal estimates $\vec{\phi}_{z}^{\mathrm{new}}$:
\begin{equation}
  \label{eq:mstepestimation}
  \vec{\phi}_{z}^{\mathrm{new}}=\pseudoinv{(\Xi_{z}\random{\Omega})}\Xi_{z}\vectorize{\random{\Lambda}},
\end{equation}
where
${\random{\Omega}=[\hadamard{(\Upsilon \random{\Theta}\Delta)}{Y};G]}$,
with matrices
${\Upsilon=\kron{\1_{n}^{T}}{I_{n}}}$,
${\Delta=\kron{I_{N}}{\1_{n}^{T}}}$,
${Y=\kron{G}{\1_{n}}}$,
${G=\kron{\1_{N}^{T}}{I_{n}}}$,
and
\[{\Xi_{z}={\diag(\sqrt{{\zeta(z_{z1};\set{P}_{\mathrm{cur}})}}I_{n},\cdots,\sqrt{{\zeta(z_{zO};\set{P}_{\mathrm{cur}})}}I_{n})}}.\]

As we can see,~\eqref{eq:mstepestimation} is a weighted Least-Squares, that uses the responsibilities as weights, ignoring those values closer to $0$.

>>>>

The main objective of the \EM{} algorithm, seen in~\cite{DempsterEtAl1977}, is to find maximum likelihood estimators for models with latent variables.

@@

For an observation ${o\in\set{O}=\{1\mathbin{:}O\}}$, we observe the input and response variables, identified as  ${\random{\vec{\theta}}_{o}}$ and ${\random{\vec{\lambda}}_{o}}$.
As~\eqref{eq:linear_cheating} gives us a multidimensional \pwa{} function, we propose to use a expansion of the model referred as \emph{mixture of switching regressions} in~\cite{QuandtRamsey1978} and \emph{mixture of linear regressions} in~\cite{FariaSoromenho2010}, which we will call \emph{mixture of affine regressions}, since our regressors have a linear term (matrices $\tilde{P}^{z}$) and a constant term (vectors $\tilde{\vec{s}}^{z}$):
% TODO(accacio): change subscript n
\begin{equation}
  \label{eq:linear_cheating_random}
  \random{\lambda}_{o}=
  \begin{cases}
    -\tilde{P}^{1}\randomvec{\theta}_{o}-\tilde{\vec{s}}^{1}&\text{with probability}\ \pi_{1} \\
    \qquad\quad \vdots&\qquad\quad \vdots\\
    -\tilde{P}^{Z}\randomvec{\theta}_{o}-\tilde{\vec{s}}^{Z}&\text{with probability}\ \pi_{Z} \\
  \end{cases}.
\end{equation}

Each couple of observed input and response variables $(    \random{\lambda}_{o}, \randomvec{\theta}_{o})$ is associated in this model to a latent unobserved random variable ${\random{z}_{o}\in\set{Z}}$ that indicates from which of the $Z$ affine regression models in~\eqref{eq:linear_cheating_random} the response variable was obtained.
The latent variable $\random{z}_{o}$ follows a categorical prior distribution, with associated probabilities ${\Pi=\{\pi_{1},\dots,\pi_{Z}\}}$:
   \begin{equation}
     \label{eq:prob_zo_equal_z}
\probability{\random{z}_{o}=z}=\pi_{z} \in [0,1], \qquad \sum_{z=1}^{Z} \pi_{z} = 1.
\end{equation}
Since $\vec{\theta}$ is our input,
we consider a non-informative improper probability density function~\cite{ChristensenEtAl2010}
% SL: BTW, maybe we should use different notations for "the probability that some discrete random variable (e.g. z_0) is equal to something, and the probability density function of a continuous random variable (e.g. theta_0 and lambda_0).

\begin{equation}
  \label{eq:theta_almost_surely}
  \probability{\randomvec{\theta}_{o}} \,\propto \,1.
\end{equation}

Given the input and latent variables, the response variable $\random{\lambda}_{o}$ is modeled as a multivariate normal random variable with probability density function
\begin{equation}
  \label{eq:multivariate_gaussian}
\probability{\randomvec{\lambda}_{o}|\randomvec{\theta}_{o},\random{z}_{o}=z; \set{P}^{z}} = \mathcal{N}(\randomvec{\lambda}_{o};f(\randomvec{\theta}_{o};\set{P}^{z}),{\Sigma^{z}}),
\end{equation}
where, following \eqref{eq:linear_cheating_random}, the mean vector is defined by
\[f(\randomvec{\theta}_{o};(P,\vec{s}))=-{P}\randomvec{\theta}_{o}-{\vec{s}},\]
and the covariance matrix ${\Sigma^{z}}$ tends to $0$.

This mixture of affine regression model is represented as a probabilistic graphical model in Fig.~\ref{fig:model}, and it corresponds to the following factorization of the complete-data (i.e., observed and latent data) likelihood, \cite{Bishop2006}:
\begin{align}\label{eq:completedataLikelihood}
  \probability{\random{\Theta},\random{\Lambda},\random{Z};\set{P}}= \prod_{o=1}^{O}\prod_{z=1}^{Z}\big[&\probability{\randomvec{\lambda}_{o}|\randomvec{\theta}_{o},\random{z}_{o}=z;\set{P}^{z}} \nonumber \\
& \times \probability{\random{z}_{o}=z}\probability{\randomvec{\theta}_{o}}\big]^{\indicator{\random{z}_{o}=z}},
\end{align}

Ideally, we would like to estimate the unknown model parameters $\set{P}$ by maximizing the log-marginal likelihood $\ln \probability{\random{\Theta},\random{\Lambda};\set{P}}$, computed by marginalizing the latent variables $\random{Z}$ in~\eqref{eq:completedataLikelihood}.
Unfortunately, this optimization problem does not admit a closed-form analytical solution.
Instead, we can exploit the latent-variable structure of the model to derive an \EM{} algorithm,~\cite{DempsterEtAl1977}, which is a convergence-guaranteed iterative algorithm ensuring at each iteration a monotonic increase of the log-marginal likelihood.

\begin{figure}[b]
  \centering
  \begin{tikzpicture}
    \draw[thick,blue,rounded corners=10pt] (-1.7,1.2) rectangle (1.7,-1.7);
    \node at (1.2,-1.3) {O};
    \node (pi) at (-2.5,.5) {$\Pi$};
    \node (phi) at (2.5,-1.) {$\set{P}$};

    \graph [edge quotes={fill=white,inner sep=1pt},
    clockwise=3,nodes={circle,draw,rotate=-60,minimum width=1cm}] {
      a/"${\randomvec{\theta}_{o}}$"[rotate=60],b/"${\random{\vec{\lambda}}_{o}}$"[rotate=60],c/"${\random{z}}_{o}$"[rotate=60];
      {a,c} ->[-latex] b;
      (pi) ->[thick,{Circle[length=3.pt]}-latex] c;
      (phi) ->[thick,{Circle[length=3.pt]}-latex] b;
      };
  \end{tikzpicture}
  \caption{Graph representation of model proposed.}\label{fig:model}
\end{figure}
% SL: observed variables are generally represented as shaded nodes in probabilistic graphical models, and latent variables as empty nodes.

@@

  Given a set of observed data $(\random{\Theta},\random{\Lambda})$, estimate the unknown latent variables $\random{Z}$ and parameter set $\set{P}$ by iteratively maximizing \todo{a function of} the complete-data log likelihood $\ln\probability{\random{\Theta},\random{\Lambda},\random{Z};\set{P}}$.
\end{problem}

Using~\eqref{eq:prob_zo_equal_z},~\eqref{eq:theta_almost_surely}, and~\eqref{eq:multivariate_gaussian} in~\eqref{eq:completedataLikelihood} yields {\color{blue}the following expression of the complete-data log-likelihood:}
\begin{equation}\label{eq:completedataLogLikelihood_complete}
\ln\probability{\random{\Theta},\random{\Lambda},\random{Z};\set{P}}=  \sum_{o=1}^{O}\sum_{z=1}^{Z}{\indicator{\random{z}_{o}=z}}
  \alpha_{zo},
\end{equation}
where ${\alpha_{zo}=\ln{\pi_{z}}+\ln{\mathcal{N}(\randomvec{\lambda}_{o};f(\randomvec{\theta}_{o};\set{P}^{z}),{\Sigma^{z}})}}$.

But as mentioned, we do not observe the complete data $(\random{\Theta},\random{\Lambda},\random{Z})$, instead we observe only $(\random{\Theta},\random{\Lambda})$ and all the information on $\random{Z}$ given these observations is carried by the posterior probabilities  ${\zeta(z_{zo};\set{P})=\probability{\random{z}_{o}=z|\randomvec{\lambda}_{o},\randomvec{\theta}_{o};\set{P}}}$, also called \emph{responsibilities}, which can be calculated as
\todo[Change Notation?]{Change Notation?}
% SL: the notation z_{zo} is a bit confusing and I am not sure it was defined. It's actually related to the fact that you use z_0 as the latent variable and $z$ as the values it can take in \mathcal{Z}. Using another variable for the latter may help clarifying.

\begin{align}
  \label{eq:responsibilities}
\zeta(z_{zo};\set{P})&=\frac{\probability{\random{z}_{o}=z}\probability{\randomvec{\lambda}_{o}|\randomvec{\theta}_{o};\set{P}^{z}}}{\sum\limits_{j=1}^{Z}\probability{\random{z}_{o}=j}\probability{\randomvec{\lambda}_{o}|\randomvec{\theta}_{o};\set{P}^{j}}}\nonumber\\
  &=\frac{\pi_{z}{\mathcal{N}(\randomvec{\lambda}_{o};f(\randomvec{\theta}_{o};\set{P}^{z}),{\Sigma^{z}})}}{\sum\limits_{j=1}^{Z}\pi_{j}\mathcal{N}(\randomvec{\lambda}_{o};f(\randomvec{\theta}_{o};\set{P}^{j}),{\Sigma^{j}})}.
\end{align}

Observe that by taking
\todo[Change Notation?]{Change Notation?}
\begin{equation}\label{eq:argmaxz}
\underset{z_{o}}{\argmax}\ {\zeta(z_{zo};\set{P})},
\end{equation}
we can get the most probable $z$-zone which generated the observation $\randomvec{\lambda}_{o}$.

So, rather than using the complete data log-likelihood~\eqref{eq:completedataLogLikelihood_complete}, which we do not have, since the latent variables are not observed by definition, the \EM{} algorithm uses its expectation with respect to the posterior calculated using a given set of parameter estimates $\set{P}_{\mathrm{cur}}$:
\begin{align}
  \label{eq:completedataLogLikelihood_expectation}
 Q\left(\set{P},\set{P}_{\mathrm{cur}}\right) &= \expectation[{\zeta(z_{zo};\set{P}_{\mathrm{cur}})}]{\ln\probability{\random{\Theta},\random{\Lambda},\random{Z};\set{P}}}
\end{align}
and since \[\expectation[{\zeta(z_{zo};\set{P}_{\mathrm{cur}})}]{\indicator{\random{z}_{o}=z}}=\zeta(z_{zo};\set{P}_{\mathrm{cur}}),\]
we can rewrite~\eqref{eq:completedataLogLikelihood_expectation}:
\begin{align}
  \label{eq:completedataLogLikelihood_expectation_developed}
Q\left(\set{P},\set{P}_{\mathrm{cur}}\right) &= \sum_{o=1}^{O}\sum_{z=1}^{Z}  \zeta(z_{zo};\set{P}_{\mathrm{cur}})\alpha_{zo},
\end{align}
\todo{\begin{remark}
For the first iteration, the model parameters must be initialized. A discussion about the initialization of the parameters is beyond the scope of this article, and we refer the reader to~\cite{Bishop2006},~\cite{DempsterEtAl1977} and their references.
\end{remark}}

Then we can find a new estimate of $\set{P}$ that maximizes $Q(\set{P},\set{P}_{\mathrm{cur}})$:
\begin{equation} \label{eq:Mstep}
  \set{P}_{\mathrm{new}}=\underset{\set{P}}{\argmax}\ Q(\set{P},\set{P}^{\mathrm{cur}}).
\end{equation}
For the next iteration of the algorithm, this new estimate $\set{P}_{\mathrm{new}}$ will be used as $\set{P}^{\mathrm{cur}}$.

With this information we can describe the \EM{} algorithm in two parts, first calculate the responsibilities $\zeta(z_{zo};\set{P}_{\mathrm{cur}})$, and then update the parameters.
Algorithm~\ref{alg:em} adapted from~\mbox{\cite[Chapter 9]{Bishop2006}} describes the steps.

@@

Here we introduce a variable ${\vec{\phi}^{z}={[\vectorize{\tilde{P}^{z}}^{T}\ {(\tilde{\vec{s}}^{z})}^{T} ]}^{T}}$ so we can calculate the new estimates of $\set{P}$ from~\eqref{eq:Mstep}. Using the KKT conditions~\cite{BoydVandenberghe2004}, we can find an optimal solution for the problem in~\eqref{eq:Mstep}, by
taking the gradients of~\eqref{eq:completedataLogLikelihood_expectation} with respect to vectors $\vec{\phi}^{z}$ and making them vanish.
Because of the multidimensional nature of the problem, some matrix operations are needed to synthesize the results.
After those operations, we have a matricial solution that yields the optimal estimates $\vec{\phi}_{z}^{\mathrm{new}}$:
\begin{equation}
  \label{eq:mstepestimation}
  \vec{\phi}_{z}^{\mathrm{new}}=\pseudoinv{(\Xi_{z}\random{\Omega})}\Xi_{z}\vectorize{\random{\Lambda}},
\end{equation}
where
${\random{\Omega}=[\hadamard{(\Upsilon \random{\Theta}\Delta)}{Y};G]}$,
with matrices
${\Upsilon=\kron{\1_{n}^{T}}{I_{n}}}$,
${\Delta=\kron{I_{N}}{\1_{n}^{T}}}$,
${Y=\kron{G}{\1_{n}}}$,
${G=\kron{\1_{N}^{T}}{I_{n}}}$,
and
\[{\Xi_{z}={\diag(\sqrt{{\zeta(z_{z1};\set{P}_{\mathrm{cur}})}}I_{n},\cdots,\sqrt{{\zeta(z_{zO};\set{P}_{\mathrm{cur}})}}I_{n})}}.\]
% SL: I find it surprising that you have square roots of the responsabilities in this equation. Are you sure it is correct?

As we can see,~\eqref{eq:mstepestimation} is the solution of a weighted Least-Squares, having the responsibilities as weights, adjusting the contribution of all observations to the regression model of index $z$.
We can see some similarities to the K-planes algorithm (see~\cite{BradleyMangasarian2000}), but \EM{} is more compromising.
Instead of affecting the observed data to a zone with 100\% of certainty (\emph{hard assignment}), \EM{} uses the responsibilities (\emph{soft assignment}) for each zone.
Then, when we solve the \textbf{M Step}, rather than using only the data assigned to update the parameters of a certain zone, we take into account all data, whose contribution to the update is weighted by their associated responsibilities.
